---
title: "MOFA INTEGRATION OF MULTI-OMICS DATA"
author: "Menna Arafat"
date:  "2024-9-25"
output: html_document
Main tutorial: https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/CLL.html
---
  

```{r}
 suppressPackageStartupMessages({ 
library("MOFA2")
library(tibble)
library(plyr)
library(dplyr)
library(tidyverse)
library(DESeq2)
library(fgsea)
library(org.Hs.eg.db)
library("pheatmap")
library("dichromat")
library("RColorBrewer")
library(ComplexHeatmap)
library(circlize)
library(stats)
library(openxlsx)
library(igraph)
library(Rgraphviz)
library(gridExtra)
 })
```

## Load precomputed model

```{r}
MOFAobject <- readRDS("output/MOFA2_TNBC.last.rds")
TNBC_mofa= MOFAobject@data
meta= MOFAobject@samples_metadata
```

## Dimensionality of the Weight matrix: 5000 features, 15 factors

```{r}
lapply(MOFAobject@expectations$W, dim)

```
##  Add sample metadata to the model

```{r}
shared_samples= MOFAobject@samples_metadata$sample
meta= data.frame(
  sample= shared_samples,
  level = ifelse(grepl("Tumor", shared_samples), "Tumor", "Normal"),
  condition = ifelse(grepl("Tumor", shared_samples), 1, 0)
)

samples_metadata(MOFAobject) <- meta
```

##visualizations

```{r}
png("output/plot_factor_cor.png",width = 2180, height = 2080, res= 600 )
plot_factor_cor(MOFAobject)
dev.off()
```

##--------------------
##plot variance explained by each factor

```{r}
library(gridExtra)
plot1 <- plot_variance_explained(MOFAobject, x = "view", max_r2 = 40)
plot2 <- plot_variance_explained(MOFAobject, plot_total = TRUE)[[2]]+ coord_flip()

png("output/plot_variance_explained_last.png", width = 7000, height = 2080, res = 600)
grid.arrange(plot1, plot2, ncol = 2)
dev.off()
```

```{r}
png("output/plot_summary_factors_scores_last.png",width = 2880, height = 1880, res= 600 )
summarise_factors(
  MOFAobject, 
  meta,
  factors = "all",
  groups = "all",
  abs = FALSE,
  return_data = FALSE
)+ theme(axis.text.x = element_text(angle = 65, hjust = 1))

dev.off()
```


##heatmap represent each factor correlation with condition (pearson corelation, p-value)

```{r}
factors <- get_factors(MOFAobject, factors = 1:15)$group1 %>% as.data.frame()
condition=  model.matrix(~ 0+meta$level) %>% as.data.frame()
names(condition)= gsub("meta\\$level", "", names(condition))
row.names(condition)= meta$sample

apply_cor= function(covariates, factors){
  corr=  cor.test(covariates, factors, method = c("pearson"))
  cor_coeff =corr$estimate %>% round(., 2)
  p_values= corr$p.value %>% round(., 2) %>% sprintf("%.2f", .)
  x <- data.frame(
    cor_coeff = cor_coeff,
    p_value = p_values,
    stringsAsFactors = FALSE)
  return(x)
}

corr_T <- mapply(function(factor_col) apply_cor(condition$Tumor, factor_col),
                 as.data.frame(factors),
                 SIMPLIFY = TRUE) %>% t() %>% as.data.frame()

corr_N <- mapply(function(factor_col) apply_cor(condition$Normal, factor_col),
                 as.data.frame(factors), 
                 SIMPLIFY = TRUE) %>% t() %>% as.data.frame()
```

##heatmap

```{r}
library(forcats)
factor= rep(paste0("Factor", 1:ncol(factors)), times = 2)

heatmap_data <- data.frame(
  factor = forcats::fct_inorder(factor),
  condition = c(rep("TNBC", ncol(factors)), rep("Normal", ncol(factors))),
  cor_coeff = c(unlist(corr_T$cor_coeff), unlist(corr_N$cor_coeff)),
  p_value = c(unlist(corr_T$p_value), unlist(corr_N$p_value)),
  stringsAsFactors = FALSE
)
```

## Create combined labels for the heatmap

```{r}
heatmap_data$label <- with(heatmap_data, 
                           paste0( sprintf("%.2f", cor_coeff), "\n",
                                   paste("(", p_value, ")") ))



plot= ggplot(heatmap_data, aes(x = factor, y = condition, fill = cor_coeff)) +
  geom_tile(color = "white") +  #' ## Draw the heatmap squares
  geom_text(aes(label = label), color = "black", size = 3) + 
  scale_fill_gradient2(low =  "white", high ="#BA55D3", midpoint = 0, 
                       limit = c(-1, 1), name = "Correlation Coeff.") +  
  labs(x = "MOFA Factors", y = "", title = "") +
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 65, hjust = 1, size = 12), 
    axis.text.y = element_text( size = 15), 
    panel.background = element_rect(fill = "white"),  
    plot.background = element_rect(fill = "white"),   
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_blank()
    
  ) 

ggsave("output/factor_condition_corr_last.png",plot, dpi = 600, width = 16, height = 9)
```


## plot factor 1, 2 to see the clustering of samples in this space

```{r}
png("output/PCA_like_plots_Factors_last.png",width = 2880, height = 1880, res= 600 )
plot_factors(MOFAobject, 
             factors = c(1,2), 
             color_by = "level",
             #' ##shape_by = "trisomy12",
             dot_size = 2.5,
             show_missing = T,
             color_name= "Groups"
)
dev.off()
```
##plot important features for each factor

```{r}
x1= plot_top_weights(MOFAobject, factors = 1, nfeatures = 10, view = 1)
x2= plot_top_weights(MOFAobject, factors = 2, nfeatures = 10, view =  1)
x3= plot_top_weights(MOFAobject, factors = 1, nfeatures = 10, view = 2)
x4= plot_top_weights(MOFAobject, factors = 2, nfeatures = 10, view =  2)
x5= plot_top_weights(MOFAobject, factors = 1, nfeatures = 10, view = 3)
x6= plot_top_weights(MOFAobject, factors = 2, nfeatures = 10, view =  3)
x7= plot_top_weights(MOFAobject, factors = 1, nfeatures = 10, view = 4)
x8= plot_top_weights(MOFAobject, factors = 2, nfeatures = 10, view =  4)
x9= plot_top_weights(MOFAobject, factors = 1, nfeatures = 10, view = 5)
x10= plot_top_weights(MOFAobject, factors = 2, nfeatures = 10, view =  5)

#x <- paste0("x", 1:10)
#cat(x, sep = ", ")
png("output/important_features_F1_F2_across_views_last.png", width = 6000, height = 9080, res = 600)
grid.arrange(x1, x2, x3, x4, x5, x6, x7, x8, x9, x10 , ncol = 2, nrow= 5)
dev.off()
```
##-----------------------------------------------------------------------------

##Heatmap for top high weigt features for each layer
##get weight matrix from MOFAobj

```{r}
MOFAobject@data_options$views

weights_F1 <- get_weights(MOFAobject, view = "Proteome", factor = 1) %>% as.data.frame()
head(weights_F1)

```

##get top weight features

```{r}
features1 = weights_F1 %>%
  arrange(desc(abs(Factor1))) %>% 
  dplyr::slice(1:50) %>% row.names()


features= c(features1 )  
```

##subset expression matrix to include only these top features

```{r}
data <- as.data.frame(TNBC_mofa[["Proteome"]])
heat_data= data[row.names(data) %in% features, ] 

row.names(heat_data)= gsub("_Methylation|_mRNA|_Proteome", "", row.names(heat_data))
head(heat_data)
```

##set annotations for heatmap from meta data

```{r}
meta$Condition= as.factor(meta$level)

ta <- HeatmapAnnotation(
  Condition = meta[, c("Condition")],
  col = list(
    Condition = c("Tumor" =   "#E3B31C", "Normal" = "#B0C4DE")
   
  ),
  annotation_height = unit(7, "mm")
)

heatmap  <- Heatmap(
  matrix = as.matrix(heat_data),
  name = "Normalized Express.data",
  col = colorRamp2(c(-3, 0, 3), c("#9370DB", "#F7F5F4","red" )),#' ##matlab::jet.colors(200),
  show_row_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = FALSE,
  top_annotation  = ta
  #' ##column_title = ""
) 
print(heatmap)
```

## Save to pdf

```{r}
png("output/heatmap_proteome_F1_50.png",width = 6000, height = 7000, res = 600)
draw(heatmap, annotation_legend_side =  "right")
dev.off()
```


##for heatmap visualization use denoised data reconconstruceted by MOFA mpde;

```{r}
factors <- get_factors(MOFAobject)$group1
weights <- get_weights(MOFAobject, view = 1)$mRNA  

reconstructed_view1 <- factors %*% t(weights)
```


##additional plots
##heatmap expression data

```{r}
plot_data_heatmap(MOFAobject, 
                  view = "mRNA",
                  factor = 2,  
                  features = 25,
                  cluster_rows = FALSE, cluster_cols = FALSE,
                  show_rownames = TRUE, show_colnames = FALSE,
                  #' ##annotation_samples = ,
                  scale = "row"
)
```

##plot heatmap with modeled data (after model estimation)

```{r}
plot_data_heatmap(MOFAobject, 
                  view = "mRNA",
                  factor = 3,  
                  features = 25,
                  denoise = TRUE, #' ##set FALSE to use original data input
                  cluster_rows = TRUE, cluster_cols = FALSE,
                  show_rownames = TRUE, show_colnames = FALSE,
                  scale = "row"
)
```

##plot high weight features

```{r}
plot_weights(MOFAobject, factors = 1, nfeatures = 20, view = 1, return_data = TRUE)

TNBC_mofa %>% names()
plot_top_weights(MOFAobject, factors = 1, nfeatures = 50, view = 5)

plot_factor(MOFAobject, 
            factors = 2, 
            color_by = "condition"
)
```

##plot data scatter

```{r}
plot_data_scatter(MOFAobject, 
                  view = "mRNA",
                  factor = 4,  
                  features = 4,
                  sign = "positive",
                  color_by = "condition"
) + labs(y="RNA expression")
```


